\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-06-27 submission/manuscript.Rmd
add

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models. We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods [Figure S3]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4]. Decision tree was more susceptible to The negative predictive impact of *Peptostreptococcus* in the decision tree model was followed by a *Lachnospiraceae* species (OTU00058) (`r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00058"))$new_auc), 2))`) [Figure 4B].

@@

As the complexity of a ML model increased [Table S1], its training times increased as well [Figure 5]. Additionally, the more hyperparameters a model had [Table S1], the longer it took to train [Figure 5]. Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

>>>>

The ML models we built using L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost decrease in interpretibility as they increase in complexity. We interpreted L1 and L2 SVM with linear kernel and L2 logistic regression using the feature weights of the trained models. We ranked the absolute weights of all the OTUs for each data-split. We calculated the median ranks of these features over the 100 data-splits. In the three linear models, OTUs that had the largest median ranks and drove the detection of SRNs belong to family *Lachnospiraceae*,  and *Ruminococcaceae* (OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609), genera *Gamella*  (OTU00426) and genera *Peptostreptococcus* (OTU00367) [Figure 3]. Some of the OTUs with the highest ranks were shared among the linear models. We explained the feature importances in non-linear models using permutation importance on the held-out test data where we randomly permuted non-correlated features individually and groups of correlated features together (see methods) to calculate the effect of permuted OTUs or group of OTUs on testing AUROC. The top 5 OTUs with the largest negative impact on testing AUROC overlapped in tree-based models [Figure 4]. Specifically, permuting *Peptostreptococcus* (OTU00367) abundances randomly, dropped the predictive performances the most for all tree-based methods [Figure S3]. Decision tree, random forest and XGBoost models' predictive performance dropped from `r paste(round(dt_median, 2))` base testing AUROC median to `r paste(round(median((dt_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, from `r paste(round(rf_median, 2))` to `r paste(round(median((rf_imp %>% filter(names=="Otu00367"))$new_auc), 2))` and from `r paste(round(xgboost_median, 2))` to `r paste(round(median((xgboost_imp %>% filter(names=="Otu00367"))$new_auc), 2))`, respectively [Figure 4].

@@

As the complexity of a ML model and the number of tuned hyperparameter settings increased [Table S1], its training times increased as well [Figure 5]. Linear models trained faster than non-linear models. L1 and L2 SVM with linear kernel and L2 logistic regression had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, a decision tree, SVM with radial basis function kernel, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

### 2019-06-27 submission/manuscript.tex
add

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
commonly see ML practices such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance or
reporting few or only the best outcomes of cross-validation. Even when
there are seperate testing sets to evaluate model performance, we
observe large differences between cross-validation and testing
performances that indicate overfitting as well as large confidence
intervals for testing performances (4, 13--19). Moreover, there is a
lack of discussion on why a particular ML model is utilized. Recently,
there is a trend towards using more complex ML models such as random
forest, extreme gradient boosting and neural networks without a
discussion on if and how much model interpretibility is necessary for
the study (11, 20--22). Black box machine learning models require
posthoc explanations to determine the importances of features in making
a prediction. These explanations can be misleading and at times
unreliable when making high-stake decisions about someone's health (23).
The lack of transparency on modeling methodology, model interpretation
and model selection negatively impact model reproducibility and
validity. We need to strive toward better machine learning practices by
(1) implementing consistent and rigourous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of model accuracy, complexity, interpretibility
and computational efficiency.

@@

We established a robust ML pipeline where we train and validate each of
the seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over the 100 randomizations and selected the best performing
hyper-parameter setting in terms of AUROC to train the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model prediction performance {[}Figure 1{]}.

@@

We reported the testing AUROC values over 100 random splits of the full
dataset which showed that depending on the data-split, the testing AUROC
value showed great variability {[}Figure 2{]}. The testing AUROC values
within each model varied 0.23 on average across the seven models. For
instance, the lowest AUROC value of the random forest model was 0.59
whereas the highest was 0.81.

>>>>

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
commonly see ML practices such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance or
reporting few or only the best outcomes of cross-validation. Even when
there are seperate testing sets to evaluate model performance, we
observe large differences between cross-validation and testing
performances that indicate overfitting as well as large confidence
intervals for testing performances (4, 13--19). Moreover, there is a
lack of discussion on why a particular ML model is utilized. Recently,
there is a trend towards using more complex ML models such as random
forest, extreme gradient boosting and neural networks without a
discussion on if and how much model interpretibility is necessary for
the study (11, 20--22). Black box machine learning models require
posthoc explanations to determine the feature importances in making a
prediction. These explanations can be misleading and at times unreliable
when making high-stake decisions about someone's health (23). The lack
of transparency on model selection and interpretation as well as flawed
modeling methods negatively impact model validity and reproducibiliyu.
We need to strive toward better machine learning practices by (1)
implementing rigourous machine learning pipelines and (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

@@

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over the 100 randomizations and selected the best performing
hyper-parameter setting in terms of AUROC to train the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model performance {[}Figure 1{]}.

@@

We reported the testing AUROC values over 100 randomizations of the
initial data-split. The results showed that depending on the data-split,
the testing AUROC value showed great variability {[}Figure 2{]}. The
testing AUROC values within each model varied 0.23 on average across the
seven models. For instance, the lowest AUROC value of the random forest
model was 0.59 whereas the highest was 0.81.
